Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-27 versions/response.v1.txt

> Associate Editor
> Editors’ comments to the author:
> All four reviewers agreed that the work is worth publishing and
> provides an important insight into the behavior of the joint
> ML/ancestral states estimation method. However, all reviewers asked
> for what amounts to major revisions to the text and presentation. You
> should especially focus on clarity and accessibility, provide the
> requested examples, and address suggestions on how the results could
> be even further improved.

Outline of revisions:
1. generalized the results to arbitrary branch lengths and not restricting the top and bottom two branches to be equal
2. expanded and clarified the exposition of the mathematics and the motivation for the Hadamard representation
3. changed discussion and plots to involve branch lengths and not fidelities

Some of the reviewers have expressed their opinion that we reduce our emphasis on mathematical development in exchange for a more numerical-optimization-based approach.
Our primary goal was to mathematically prove inconsistency of joint inference, which is what we have done.
As pointed out by reviewer 4, this answers an open question, which has special relevance given the recent work of Sagulenko et al.
We provide some numerical results to confirm our derivations and to provide a comparison.

%das: added
This work is not the final word on the joint inference question from an applied perspective, as we now emphasize with an expanded Discussion section as well as the addition of these lines:
"Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
Further study is needed to determine the impact of these results on problems that are of interest to practitioners."

Even if we were to greatly expand the sections doing numerical optimization, these results would only be for the binary symmetric model, which is never applied in practice.
An investigation of these aspects deserves another paper.
We have added clarification of these points to the discussion:
"All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Examples with larger trees and more realistic mutation models fall within the scope of an extension to applied models.
Further study is needed to determine the impact of these results on problems that are of interest to practitioners."

@@

> Finally, please pay VERY careful attention to improving your
> literature review (reviewer 4) and making a very clear case about how
> your work IMPROVES and EXTENDS previous results. And, please, do get
> rid of the Farris zone. The name "Farris" should not be attached to
> results that have to do with ML estimation, given the long and
> torturous efforts of cladists to undermine the entire line of inquiry.

We are grateful to reviewer 4 for pointing us to literature of which we were not aware.
We have greatly expanded on the discussion in Goldman (1990) about the limitations of joint optimization of phylogenetic parameters and ancestral states (the newly added fourth paragraph of the Discussion section), and have chosen to use his terminology of "integrated likelihood" to avoid confusion.
The previous work concerns an example where branch lengths are fixed and equal, while here we show cases where inconsistency arises when branch lengths are treated as unknown parameters.
Additionally, we have placed joint optimization and maximum likelihood on level footing by viewing them both as optimization procedures over topologies and branch lengths.
Goldman (1990) does not compare maximum integrated likelihood and maximum relative likelihood as the latter assumes equal branch lengths while the former does not.
We have added this to our discussion as well.

We have complied with the request regarding the use of "Farris tree," despite it going against current practice.
Indeed, Felsenstein's book has a subsection entitled "The Farris zone" spanning two pages (pp 120-121), a figure entitled "A tree in the Farris zone.", and a table with a caption "Properties of the Felsenstein and Farris zones...".
The term "Farris zone" has almost 7 times as many citations as the term "inverse Felsenstein zone" on Google Scholar.

@@

> Reviewer: 1

> Comments to the Author
> The authors show that when ancestral states are considered as
> parameters, there are zones where topologies and edge-lengths are
> inconsistently estimated. As is evidenced by any ML package, it is
> widely accepted that ancestral states should not be parameters; the
> large number of parameters suggests the approach will be problematic.
> Still, to the best of my knowledge, it has not been shown that
> treating ancestral states as parameters actually leads to poor
> performance, or how poor. Moreover, the authors point to Sagulenko et
> al (2018) - now published in Virus Evolution - as an example
> implementation. I found the paper interesting and I believe a more
> succinct presentation will be of interest.

> A main concern with the paper is that the shaded regions of
> topological inconsistency or poor estimation in Figs 2-4, are only
> subsets of the actual regions where topological inconsistency or poor
> estimation occur; and yet determining the actual regions seems
> feasible. The contrast between Fig 5 and 4 illustrates how far off the
> subsets are in giving the actual regions where estimation is
> problematic. The white region in Fig 5 is the actual region where hat
> w=1 whereas the shaded region of Fig 3 is the subset of the actual
> region that the authors were able to determine through analytical
> arguments. For most readers, out of Fig 3 and Fig 5, only Fig 5 is of
> interest.

As described above, our intent is to mathematically establish inconsistency and poor estimation.
This must be done with analytic arguments rather than optimization.
We could have only shown this bad behavior in a very small area to establish that it happens.
In contrast to this reviewer, we believe that Fig 5 is optional to this line of argument, and we provide it just for interest.
A true investigation into the practical implications of our work will require another paper.
We have clarified this in the Discussion section with:
"Our aim is to analytically demonstrate joint inference is inconsistent, and, while this indicates an issue with this inference procedure, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in Lemma 3 provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation."

@@

> I know that it's not feasible to explicitly compute (analytically)
> these likelihoods for bigger trees. But one of the main motivations
> and applications for tools like TreeTime is surely that the number of
> sequences grows (not the number of sites, which is fixed by the viral
> sequence length in the case of viruses). What happens when the number
> of tips grows? Are there more opportunities for the kind of issue that
> gives the wrong value for w? Many influenza trees have many 0 (or near
> 0) internal branch lengths, and it would be interesting to understand
> the implications for the ancestral state reconstructions (which in
> turn are used to infer spatial spread, transmission etc).

%EM David-- does it seem like we could extend these results to bigger trees containing a 4 taxon tree?
%das: I wasn't able to come up with anything interesting for this.

> Some detailed comments and questions:
> – line 85 some cut/paste or sentence issue

We have added the definition of a "generating probability" to ease confusion in this sentence:
"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."

@@

> - line 97 below (2) define A^p

We have clarified this with an explicit definition of A^p where we introduce A, at the beginning of the section on Phylogenetic maximum likelihood.

> - l106: why is the functional form of (3) determined by inequalites;
> what are they?

We have made this more precise and directly referenced the inequalities:
"In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state split, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}_\nCols$ for a fixed $(\tau,t)$, then $(\hat{\tau}_\nCols,\hat{t}_\nCols)$ given $\hat{\fullAncestralStates}_\nCols$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}."

@@

> - l 118-124: clarify are tau1 and tau2 Farris and Felsenstein?

Thank you for highlighting this omission.
You are correct, and we have added parenthetical notes to remind the reader where appropriate.

@@

> L 132 clarify “equivalent but different notion” – different from what?

We have reformulated this sentence to avoid the comparison:
"We parameterize the branches of these trees not with the standard notion of branch length in terms of number of substitutions per site, but with an alternate formulation called "fidelity.""

@@

> - line 146 “seeing no change along a branch more likely increases the
> likelihood by introducing a term of (1 + θ) instead of (1 − θ)” – I
> don't understand this

TODO: this is changed, but I don't see this being a lot clearer.
%das: I believe this is clear now; lmk
We have clarified the intuition in this sentence:
"For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth."

@@

> line 185++ and figure5: true value of w?

We have added clarifying sentence:
"As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$."

@@

> L 334 missing “the”

Fixed via simpler language.
We removed the sentence following the semicolon entirely, as we feel it did not add anything:
"Since we have a finite character alphabet, for a given column $i$ there are a finite number of possible assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$."

@@

> L332++ I got confused about the complements and factors of 2, though I
> see the point that the site split has symmetry. As it's fundamental to
> the rest of the work, a more clear statement of where the 2 comes from
> in the second eq below 358 would be very helpful.

We have now made this much more explicit with three additional display equations spelling out the intermediate steps.

> L354 why does psi take the complement of site patterns? Just above it
> is defined as a random variable psi (Y) where Y is a site pattern

We apologize for the confusion, as this was not as clear as it should have been.
We have revised to:
"The mapping $\patternToSplit$ operates by returning the tips labeled as $1$ in a site pattern to obtain a site split in $\mathcal{P}(\siteSplitSet)$ if the set of tips labeled $1$ is not in $\mathcal{P}(\siteSplitSet)$.
The mapping $\ancestralToSplit$ is defined by whether the tip states have their complements taken or not: if the set of tips labeled $1$ in $\alignmentColumn$ is in $\mathcal{P}(\siteSplitSet)$, $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ is the set of tips labeled $1$ in $\ancestralStateColumn$; otherwise, the set of tips labeled $1$ in $\overline{\alignmentColumn}$ necessarily is in $\mathcal{P}(\siteSplitSet)$."

Additionally, we have explicitly defined psi(Y) and xi(Y,H) prior to this clarifying sentence.

@@

> What are the elements of the sets eta(tau, t)?

We removed the ordered list of sets as it did not add to any understanding of the concept and now refer to each of these sets and their relevance directly.

> Why is xi surjective? I don't understand the equation before line 359,
> and then I don't understand what “such a choice” is in line 359

We have made this explicit now, with two additional intermediate steps and an explicit explanation in terms of collapsing terms:
"For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}\neq 1$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective."

@@

> line 364 define p(e)

Thanks for reminding us to define p(e), which is now described as "where $p(e)$ is the probability of a character change along edge $e$."

> line 366  I don't understand (8) but perhaps that's just background I
> don't have – a quick explanation or reference would be handy
> What's a 'generating probability' ? Is that the fidelity? Or a
> generator?

We have updated the text to explicitly define "generating probability."
"As data becomes plentiful, we use the Hadamard transform (see (9) in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form."

Background for equation (8) can be found in Section 8.6 in the Semple and Steel text.
This is referenced at the beginning of the section:
"We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}."

We do not believe an exhaustive explanation would provide any benefit to the reader that they could not get from referring to the text.

@@

> Comments to the Author
> Shaw and Matsen show that a joint maximum-likelihood inference of
> phylogenies and ancestral states is not consistent when branch lengths
> are long. We have recently proposed such a joint estimation for
> sequences that recently diverged and are hence similar (Sagulenko et
> al). The results presented here are anticipated in Sagulenko et al
> "For more divergent sequences, however, subleading states of internal
> nodes make a substantial contribution and the iterative optimization
> will underestimate the branch lengths."
> and
> "The iterative optimization procedures are not expected to be accurate
> for trees were many sites are saturated. In scenarios with extensive
> uncertainty of ancestral states and tree topology, convergence of the
> iterative steps cannot be guaranteed."

> In this light, it should be pointed out that joint estimation
> procedure was explicitly suggested as an approximation and hence never
> expected to be consistent in a strict sense. Furthermore, it was
> suggested as an approximation when branch lengths are short -- the
> opposite of the problematic regions identified here.
> But systematic analysis of the behavior of this approximation is
> useful and I believe the work is correct (but haven't checked the
> proofs). The work by Shaw and Matsen suggests the radius of
> convergence of this procedure is somewhere around branch length of
> order 10-20%, which is consistent with results reported by Sagulenko
> et al.

Thank you for your thoughts, and we have worked hard to not oversell our results.
The TreeTime paper does present joint optimization as an approximate approach, and described that it is for short branch lengths.
However, in our assessment the word "approximate" doesn't completely describe a situation in which there is significant bias.
The TreeTime paper says that its use is appropriate when "ancestral sequence inference has little ambiguity" but even in these cases the bias in branch lengths persists.

TODO we should make some changes to at least say that we've further clarified the implications of our results.
%das: I believe this is done
We clarify our aim and warnings from Sagulenko et al. (2017) in the Introduction:
"Although the software described in \citet{Sagulenko2018-xl} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference."

Additionally, we refer to how our work extends that from Sagulenko et al. (2017) in the Discussion:
"The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3)."

@@

> At the same time, these results are not particularly surprising.
> Assigning ML ancestral states will always result in shorter branches.
> For short branches, this correction is linear in branch length, for
> longer branches higher order effects start to contribute. If the
> typical branch length around a bifurcating node is x<<1, a fraction ~x
> of the sites mutated along one of the three adjacent branches, while a
> fraction ~x^2 mutated on two branches and not the third. The latter
> case is ignored in this joint inference procedure. Not having to sum
> over these subleading states is what makes joint inference faster, but
> such truncation will first lead to inaccuracies and at longer
> distances to qualitatively different solutions (like Taylor expanding
> 1/(1-x) ).
>
> There are other obvious cases where joint inference will fail. For
> sequences with a heavily biased in nucleotide composition (say biased
> towards A). For sufficiently long terminal branches, all ancestral
> states will be estimated as all 'A' and subsequent branch length
> estimate will shrink all internal branches to zero. This happens for
> the very same reason that the most likely outcome of multiple flips of
> a biased coin is all heads -- a very atypical yet most likely outcome.
>
> Currently, the work largely shows that areas of inconsistencies exist.
> It would be useful if the authors could present Taylor expansions for
> the ML branch length estimates around x=y=w=1. This would more
> explicitly show the inaccuracies accumulate with increasing branch
> length and how the inaccuracies depend on asymmetries in the tree. To
> me, this would be much more useful than to know that joint inference
> will fail if branches are long.

This is an interesting suggestion but is different than our goal, which was to establish inconsistency.
Moreover, in the revised manuscript we see cases where assigning ML ancestral states results in longer branches, not shorter ones (Lemma 3).
We believe the inaccuracies have more to do with the specific choice of which ancestral state is maximal than asymmetries in the tree:
"This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods})."

Asymmetries can drive these particular choices, but Fig. 3 in the revised manuscript shows that almost half of the parameter space results in an interior branch length being estimated as zero.
We do not think providing an argument about accumulating inaccuracies would elucidate why such starkly-defined regions of inconsistency exist.
For reasonable parameter settings, one could investigate the error accumulation given fixed ancestral states, but this was not our goal and will require another paper.

TODO describe why Taylor isn't a straightforward thing.
TODO refer to idea of applied paper again?
%das: I believe these are done as well

@@

> Specific comments:
>  - when y=x=w, the two topologies are indistinguishable other than for
>  the labeling of the leaves. What happens in Fig 2 on the diagonal?
>  Does it degenerate into a star?

It is true when y=x=w we are unable to distinguish the topologies apart from the labeling, though the labeling is what distinguishes the topologies.
As shown in Goldman (1990), with all branch lengths fixed and equal the joint inference procedure degenerates into parsimony, which has well known inconsistency issues.
We have added discussion of this point:
"The case of joint inference of a phylogenetic likelihood is discussed in \citet{Goldman1990-dk}.
There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general."

@@

>  - line 50: overestimates branch lengths? It overestimates fidelity ->
>  underestimates branch length.  As expected.  Again line 249. Please
>  check. The back and forth between fidelity and branch length can be
>  confusing.

### 2018-07-30 appendix.tex
hedging and some tense changes
A handful of appendix changes.

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
\end{lemma}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
We focus on $x_1$.
The marginal log-likelihood proportional to $x_1$ can be written as
\begin{align*}
\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\end{align*}
where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example,
\[
c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
\end{align*}
and focus on a single term
\[
f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
\]
We see
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
\]
which greater than or equal to zero as it is a probability.
Thus, regardless of $a_{\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 \ge 0$.
The remaining variables follow similar arguments.

@@

As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
As we are interested in a lower bound, assume $\boldsymbol\xi=\boldsymbol\xi_0$, i.e., $\emptyset$ is the maximal ancestral state split for all site splits.
In this case, the conditional log-likelihood for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same conditional log-likelihood for both topologies.

>>>>

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
\]
in general.

@@

\begin{lemma}
\label{lemma:bounds}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that $\emptyset$ is the maximal ancestral state split for all site splits given $\tau=\tau_1$.
\end{lemma}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We will show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
We focus on $x_1$.
%EM I think you mean the divergence term as a function of x_1 is proportional to...
The marginal log-likelihood proportional to $x_1$ can be written as
\begin{align*}
%EM This is a little funky-- I think this is a definition of D(x_1)?
\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\end{align*}
where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example,
\[
c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
\end{align*}
and focus on a single term
\[
f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
\]
We see
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Thus, regardless of $a_{\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 \ge 0$.
The remaining variables follow similar arguments.

@@

%das: @EM did you say you wanted to get rid of these paragraphs? we never show there is a case where Fels > InvFels, just where it's equal...
As the region in Fig.~\ref{fig:incons-analytic} gives the values of $x^*$ and $y^*$ where $\hat{w}$ is guaranteed to be one, we converge on a multifurcating topology in these cases.
We show this is a type of inconsistency in topology estimation by constructing a lower bound on the likelihood for the Felsenstein topology $\tau_2$ and demonstrating it is equal to the likelihood given the InvFels topology $\tau_1$ in the demarcated region of Fig.~\ref{fig:incons-analytic}.
As we are interested in a lower bound, assume $\boldsymbol\xi=\boldsymbol\xi_0$, i.e., $\emptyset$ is the maximal ancestral state split for all site splits.
In this case, the conditional log-likelihood for $\tau_2$ is composed of the same functional forms given in Table~\ref{tab:likelihoods-restricted}, as assuming $\emptyset$ for the ancestral state split results in the same conditional log-likelihood for both topologies.
